---
title: "The Leslie Matrix - revisited"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

A population of fish we're interested in has four life stages: eggs (E), fry (F), juvenile (J), breeding adult (A). You are told that annually:

- Each breeding adult will produce, on average, 600 eggs
- 5% of eggs will survive to become fry
- 20% of fry will survive to become juveniles
- 16% of juveniles will survive to become adults
- Adult survival rate year to year is 90%

Recall the Leslie matrix from EDS 212. Matrix multiplication (using the dot product `%*%` of the matrix with the populations at each life stage) yields the population structure at the following year. 

### Create the Leslie matrix:

```{r}
# Make the Leslie Matrix:
#this makes a matrix using two linear equations
fish_leslie <- matrix(c(0, 0, 0, 600, 0.05, 0, 0, 0, 0, 0.20, 0, 0, 0, 0, 0.16, 0.9), nrow = 4, ncol = 4, byrow = TRUE)

# Check it out: 
fish_leslie
```

The initial population structure is given by: 
```{r}
# Initial population structure (Roe, Fry, Juvenile, Adult): 
fish_year0 <- c(5000, 8000, 600, 400)
```


Add line-by-line comments wheres prompted in the code below, explaining what each is doing. Make sure that you are talking with your neighbors about what is really being created at each step. 

```{r}
# Model projections

# ------ WHAT HAPPENS HERE? ------ #
# ------ comment line-by-line ---- #

# The line below...
#creates a sequence from 0 to 8 by increments of 1; 9 rows and 4 columns
time_yr <- seq(from = 0, to = 8, by = 1)

# The line below...
#creates empty matrix where fish year=col and time passed=row
proj_year <- matrix(ncol = length(fish_year0), nrow = length(time_yr))

# The line below...
#this sets up the first year(year 0)
#[row, column]; empty [1, ]means go through all the columns
proj_year[1, ] <- fish_year0

# The for loop below...
#this populates the matrix downwards and takes the dot product of two linear equations and the previous year
# : means from:to, so from 2 to year 9 (really t=8 but we start at 0), since the first year was populated in line code above. The length (time_yr) makes it easier to change value of years
for (i in 2:length(time_yr)) {
  proj_year[i,] <- fish_leslie %*% proj_year[i-1,]
}


# The rest is wrangling & visualization (run to check it out):
#this names the columns
colnames(proj_year) <- c("eggs", "fry", "juvenile", "adult")

#Good note - you can select certain parts of the code and run that only. Good to run code line by line as practice.
#this takes your value and melts it into a longer/narrower table format, which is easier for coding
proj_df <- data.frame(time_yr, proj_year) %>% 
  pivot_longer(cols = -time_yr, names_to = "lifestage", values_to = "stage_population")

#this uses aes(color) to group lifestages by color
ggplot(data = proj_df, aes(x = time_yr, y = stage_population)) +
  geom_line(aes(color = lifestage)) +
  scale_y_log10()

```

Test different combinations of survival / reproduction and starting population structures to see what happens! 
